---
title: "Bayesian Optimal Interval Design"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{Bayesian Optimal Interval Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
The Bayesian Optimal Interval (BOIN) design was introduced by @liu_bayesian_2015.
It is one of a series of dose-finding trial designs that works by partitioning the probability of toxicity into a set of intervals.
These designs make dose-selection decisions that are determined by the interval in which the probability of toxicity for the current dose is believed to reside.


# Summary of the BOIN Design
TODO


# Implementation in `escalation`
TODO

```{r, message=FALSE}
library(escalation)
```


## Dose paths
TODO

For more information on working with dose-paths, refer to the [dose-paths vignette](https://brockk.github.io/escalation/articles/DosePaths.html).


## Simulation
TODO

For more information on running dose-finding simulations, refer to the [simulation vignette](https://brockk.github.io/escalation/articles/Simulation.html).


# References
