---
title: "EffTox, Wages & Tait, and BOIN12 - Efficacy & Toxicity Designs"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{EffTox, Wages & Tait, and BOIN12 - Efficacy & Toxicity Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 5
)
```

# Introduction
Seamless phase I/II designs extend the typical phase I or _maximum tolerated dose_-seeking approach by incorporating efficacy outcomes into the dose selection decision.
Researchers will be motivated to escalate dose whilst there is a lack of efficacy and de-escalate under the presence of excess toxicity.

One of the earliest designs in this space was EffTox by @thall2004dose.
Their design uses logit models for the probability of toxicity and efficacy at the research doses, and allows the two events to be correlated in a joint model.
@wages2015seamless introduced a multi-stage design that uses response-adaptive randomisation to explore the doses.
Most recently, @lin2020boin12 introduced a generalisation of the BOIN Bayesian interval design for efficacy and toxicity outcomes called BOIN12.
These designs are implemented in `escalation` and demonstrated below.

First, we must learn how to specify efficacy and toxicity outcomes in `escalation`. 
We do that in the next section.

# Outcome nomenclature
In a joint phase I/II trial, where we have coincident efficacy and toxicity outcomes, we present outcomes in the individual patients using the letters:

* `T` to show that toxicity without efficacy occurred in a patient;
* `E` to show that efficacy without toxicity occurred in a patient;
* `N` to show that neither occurred;
* `B` to show that both occurred.

These outcome letters are strewn behind dose-levels to show the outcomes of patients in cohorts.
To show that a cohort a three patients was given dose 2, with the first experiencing toxicity without efficacy, the second experiencing efficacy without toxicity, and the third experiencing both, we would use the outcome string:

```{r}
outcomes <- '2TEB'
```

There are many more examples below.


# EffTox
Thall & Cook [@thall2004dose] introduced the _EffTox_ design for dose-finding clinical trials where both efficacy and toxicity events guide dose selection decisions.
We provide a brief recap of EffTox here, but full details are given in [@thall2004dose; @thall2006; @thall2014].

For doses $\boldsymbol{y} = (y_1, ..., y_n)$, the authors define codified doses $(x_1, ..., x_n)$ using the transform 

$x_i = \log{y_i} - \sum_{j=1}^n \frac{\log{y_j}}{n}$

The codified doses are used as the explanatory variables in logit models for the marginal probabilities of toxicity and efficacy:

$\text{logit } \pi_T = \alpha + \beta x$

$\text{logit } \pi_E = \gamma + \zeta x + \eta x^2$

Let $(Y_j, Z_j)$ be random variables each taking values $\{0, 1\}$ representing the presence of efficacy and toxicity in patient $j$. 

Then the efficacy and toxicity events are associated by the joint probability function

$Pr(Y = a, Z = b) = \pi_{a,b}(\pi_E, \pi_T) = (\pi_E)^a (1-\pi_E)^{1-a} (\pi_T)^b (1-\pi_T)^{1-b} + (-1)^{a+b} (\pi_E) (1-\pi_E) (\pi_T) (1-\pi_T) \frac{e^\psi-1}{e^\psi+1}$.

Normal priors are specified for the elements of the parameter vector $\boldsymbol{\theta} = (\alpha, \beta, \gamma, \zeta, \eta, \psi)$.

At each dose update decision, the dose $x$ is acceptable if

$\text{Pr}\left\{ \pi_T(x, \boldsymbol{\theta}) < \overline{\pi}_T | \text{data} \right\} > p_T$

and

$\text{Pr}\left\{ \pi_E(x, \boldsymbol{\theta}) > \underline{\pi}_E | \text{data} \right\} > p_E$

and is no more than than one position below the lowest dose-level given and no more than one position above the highest dose-level given.
The net effect of these last two criteria is that untried doses may not be skipped in escalation or de-escalation.
$\underline{\pi}_E, p_E, \overline{\pi}_T, p_T$ are provided by the user as the trial scenario dictates.

The utility of dose $x$, with efficacy $\pi_E(x, \boldsymbol{\theta})$ and toxicity $\pi_T(x, \boldsymbol{\theta})$ is 

$u(\pi_E, \pi_T) = 1 - \left( \left(\frac{1-\pi_E}{1-\pi_{1,E}^*}\right)^p + \left(\frac{\pi_T}{\pi_{2,T}^*}\right)^p \right) ^ \frac{1}{p}$

where $p$ is calculated to intersect the points $(\pi_{1,E}^*, 0)$, $(1, \pi_{2,T}^*)$ and $(\pi_{3,E}^*, \pi_{3,T}^*)$ in the efficacy-toxicity plain.
We refer to these as _hinge points_ but that is not common nomenclature.

At the dose selection decision, the dose-level from the acceptable set with maximal utility is selected to be given to the next patient or cohort.
If there are no acceptable doses, the trial stops and no dose is recommended.

There are several published EffTox examples, including explanations and tips on parameter choices  [@thall2004dose; @thall2006; @thall2014].

# Wages & Tait
The design of @wages2015seamless uses latent CRM models to estimate the probabilities of toxicity and efficacy, and two stages to identify the recommended dose.
In the first stage, the design sequentially seeks the dose with probability of toxicity nearest to an upper toxicity threshold, like the classic CRM approach.
In the second stage, the design adaptively randomises to the permissible doses, i.e. those with probability of toxicity not considered statistically too toxic, with greater randomisation weight given to those doses with higher estimated efficacy rates.
The design uses a toxicity skeleton, like CRM, and several competing efficacy skeletons to accommodate that efficacy may not be monotonically increasing in dose.

# BOIN12
The Bayesian Optimal Interval (BOIN) design was introduced by @liu_bayesian_2015.
It is one of a series of dose-finding trial designs that works by partitioning the probability of toxicity into a set of intervals.
These designs make dose-selection decisions that are determined by the interval in which the probability of toxicity for the current dose is believed to reside.
The BOIN12 generalisation applies similar logic to co-primary binary toxicity and efficacy outcomes.

# Implementation in `escalation`
To demonstrate the methods, let us consider the recommended dose after observing outcomes `1NNN 2EBT` seeking a dose no more toxic than 0.3 with an associated efficacy probability of at least 0.5.

In each design, there will be many parameters that need to be set because these designs are comparatively complex.
Ordinarily, parameter choices would be justified in light of the prevailing clinical scenario, and evaluated by simulation.
For the purposes of illustration in this document, however, we are going to proceed with typical or general values.

```{r, message=FALSE}
library(escalation)

outcomes <- "1NNN 2EBT"
```


## EffTox
For simplicity, we will use the parametersiation specified in an example investigating five doses in @thall2014:

```{r}
p <- trialr::efftox_priors(
  alpha_mean = -7.9593, alpha_sd = 3.5487,
  beta_mean = 1.5482, beta_sd = 3.5018,
  gamma_mean = 0.7367, gamma_sd = 2.5423,
  zeta_mean = 3.4181, zeta_sd = 2.4406,
  eta_mean = 0, eta_sd = 0.2,
  psi_mean = 0, psi_sd = 1
)
real_doses = c(1.0, 2.0, 4.0, 6.6, 10.0)
num_doses <- length(real_doses)

et_model <- get_trialr_efftox(
  real_doses = real_doses,
  efficacy_hurdle = 0.5, toxicity_hurdle = 0.3,
  p_e = 0.1, p_t = 0.1,
  eff0 = 0.5, tox1 = 0.65,
  eff_star = 0.7, tox_star = 0.25,
  priors = p, 
  # Low MCMC sample size purely for speed in vignette:
  iter = 1000, chains = 1, seed = 2020
)
```

```{r}
et_fit <- et_model %>% fit(outcomes)
```

```{r}
et_fit
```

```{r}
recommended_dose(et_fit)
```

```{r}
continue(et_fit)
```

```{r}
dose_admissible(et_fit)
```


## Wages & Tait
For simplicity we will use the parametersiation specified in @wages2015seamless, tweaked slightly to reflect the dose-levels and targets specified in the EffTox example above.


```{r}
tox_skeleton = c(0.08, 0.15, 0.22, 0.29, 0.36)

eff_skeletons = matrix(nrow = 9, ncol = num_doses)
eff_skeletons[1,] <- c(0.60, 0.50, 0.40, 0.30, 0.20)
eff_skeletons[2,] <- c(0.50, 0.60, 0.50, 0.40, 0.30)
eff_skeletons[3,] <- c(0.40, 0.50, 0.60, 0.50, 0.40)
eff_skeletons[4,] <- c(0.30, 0.40, 0.50, 0.60, 0.50)
eff_skeletons[5,] <- c(0.20, 0.30, 0.40, 0.50, 0.60)
eff_skeletons[6,] <- c(0.30, 0.40, 0.50, 0.60, 0.60)
eff_skeletons[7,] <- c(0.40, 0.50, 0.60, 0.60, 0.60)
eff_skeletons[8,] <- c(0.50, 0.60, 0.60, 0.60, 0.60)
eff_skeletons[9,] <- c(0.60, 0.60, 0.60, 0.60, 0.60)
eff_skeleton_weights = rep(1, nrow(eff_skeletons))

wt_model <- get_wages_and_tait(
  tox_skeleton = tox_skeleton,
  eff_skeletons = eff_skeletons,
  tox_limit = 0.3, eff_limit = 0.5,
  num_randomise = 20
)
```

The toxicity skeleton specified above plays exactly the equivalent role as seen in the CRM design.
There are several efficacy skeletons to allow that efficacy may not be monotonic in dose.
We use efficacy skeletons to reflect that: i) each dose may be the maximum, with monotonic increase and subsequent decrease relative to that dose; and ii) that each dose may be a plateau point, with equal efficacy at higher doses. 
In general, this generates $2n - 1$ efficacy skeletons in a trial of $n$ doses.
The probabilities in the efficacy skeletons are less important than the shape the skeleton conveys because the average height of the dose-efficacy curve will be adjusted in Bayesian updating.

We fit the model to the outcomes:

```{r}
wt_fit <- wt_model %>% fit(outcomes)
```

and see that a dose is advised for the next cohort:

```{r}
recommended_dose(wt_fit)
```

```{r}
continue(wt_fit)
```

Note that this design differs from EffTox in which doses it infers to be admissible.
Different designs behave in different ways!

```{r}
dose_admissible(wt_fit)
```

Uncommonly seen in dose-finding designs, this design is a randomising design:

```{r}
is_randomising(wt_fit)
```

Doses are assigned to patients in the first trial stage in a non-randomising way (i.e. the same outcomes fit to the same design will yield the same recommended dose).
In contrast, doses are randomised in the second trial stage correlated to the modelled efficacy probability at each dose.


## BOIN12
For illustration, we use the parameterisation presented and justified in @liu_bayesian_2015.

```{r}
b_model <- get_boin12(
  num_doses = 5, phi_t = 0.3, phi_e = 0.5, u2 = 40, u3 = 60, n_star = 6
)
```

We fit the model to outcomes using the usual method:

```{r}
b_fit <- b_model %>% fit(outcomes)
```

And we can obtain the usual information as presented in many places in these vignettes:

```{r}
recommended_dose(b_fit)
```

```{r}
continue(b_fit)
```

```{r}
dose_admissible(b_fit)
```

Unlike EffTox and Wages & Tait, the BOIN12 design performs an additional step at the final dose recommendation to select the dose that is finally recommended.
This same approach is used in BOIN and mTIP2, for instance.
Read on.

### Final dose selection
BOIN12, like some other designs, selects the final dose differently to how it selects doses mid-trial.
To achieve this in `escalation`, we need an extra selector that will kick-in when the parent selector(s) have selected a non-NA dose but expressed `continue == FALSE`, i.e. signaled _the trial ends now but we are interested in a dose_.
If used, it will almost surely come last in the selector chain:

```{r}
b_model2 <- get_boin12(
  num_doses = 5, phi_t = 0.3, phi_e = 0.5, u2 = 40, u3 = 60, n_star = 6
) %>%
  stop_at_n(n = 12) %>%
  select_boin12_obd()

outcomes <- '1NNN 2NTN 2NNN 3BEN'
b_model2 %>% fit(outcomes) %>% recommended_dose()
```

Note we have forced the maximum sample size to 12 and provided outcomes for 12 patients so that the logic of `select_boin12_obd` is invoked.


In the above example, `stop_at_n` stopped the trial because the threshold sample size was met, and the underlying algorithm identified that at least one dose was worthy of selection.
At this juncture, `select_boin12_obd` took over and applied the method described by the authors.
Whilst the underlying selector(s) were busy conducting the trial (`continue == TRUE`), `select_boin12_obd` kept silent.

# Dose paths
As with all designs in `escalation`, we can use dose-paths to exhaustively calculate model advice for all possible future cohorts.
However, first, a note of warning.

In phase 1/2 trials, there are four outcomes each patient may experience (listed above), so there are 10 distinct combinations of outcomes that a cohort of two patients may yield (n.b. we are counting combinations rather than permutations), and 20 combinations that a cohort of three may yield.
This means that the size of a dose-path tree grows very quickly, so we must be modest in our aspirations here lest we be overwhelmed with information!

For example, after observing `1NN` in an initial cohort of two patients, we will reproduce the dose-advice for the next two cohorts of two patients (i.e. 100 paths).
For comparison, we will specify that the next dose will be 2.
This starts the designs from a common place.
Naturally, this option can be omitted and the design would start from whatever dose it recommended after seeing outcomes `1NN`.

To visualise paths for the EffTox design, we run:

```{r}
cohort_sizes <- c(2, 2)
outcomes <- "1NN"

et_paths <- et_model %>% 
  get_dose_paths(
    cohort_sizes = cohort_sizes, 
    previous_outcomes = outcomes,
    next_dose = 2
  )

graph_paths(et_paths)
```

Had there been no previous outcomes (i.e. at the start of a new trial), we could have omitted the `previous_outcomes` parameter.

We can visualise the equivalent using the Wages & Tait design:

```{r}
wt_paths <- wt_model %>% 
  get_dose_paths(
    cohort_sizes = cohort_sizes, 
    previous_outcomes = outcomes,
    next_dose = 2
  )

graph_paths(wt_paths)
```

and the BOIN12 design:

```{r}
b_paths <- b_model %>% 
  get_dose_paths(
    cohort_sizes = cohort_sizes, 
    previous_outcomes = outcomes,
    next_dose = 2
  )

graph_paths(b_paths)
```

As we might expect, we see distinct differences in the dose decisions by design.
EffTox will escalate in this example much more readily.

For more information on working with dose-paths, refer to the dose-paths vignette.


# Simulation
We can simulate trials using all designs in `escalation`.
Computation using Bayesian designs can be computationally costly so for the purposes of illustration in this vignette, we will simulate just twenty trials per design at the following true toxicity and efficacy curves:

```{r}
n_sim <- 20
true_prob_tox = c(0.02, 0.12, 0.17, 0.38, 0.55)
true_prob_eff = c(0.25, 0.27, 0.52, 0.54, 0.54)
```

To promote comparability, we will give each design just 12 patients (again, just for illustration) and prevent each design from skipping doses in escalation.

To simulate under the EffTox design, we run:

```{r}
et_model <- get_trialr_efftox(
  real_doses = real_doses,
  efficacy_hurdle = 0.5, toxicity_hurdle = 0.3,
  p_e = 0.1, p_t = 0.1,
  eff0 = 0.5, tox1 = 0.65,
  eff_star = 0.7, tox_star = 0.25,
  priors = p, 
  # Low MCMC sample size purely for speed in vignette:
  iter = 2000, chains = 1, 
) %>% 
  dont_skip_doses() %>% 
  stop_at_n(n = 12)

set.seed(2025)
et_sims <- et_model %>% 
  simulate_trials(
    num_sims = n_sim, 
    true_prob_tox = true_prob_tox,
    true_prob_eff = true_prob_tox
  )
et_sims
```

And under the Wages & Tait design:

```{r}
wt_model <- get_wages_and_tait(
  tox_skeleton = tox_skeleton,
  eff_skeletons = eff_skeletons,
  tox_limit = 0.3, eff_limit = 0.5,
  num_randomise = 20
) %>% 
  dont_skip_doses() %>% 
  stop_at_n(n = 12)

set.seed(2025)
wt_sims <- wt_model %>% 
  simulate_trials(
    num_sims = n_sim, 
    true_prob_tox = true_prob_tox,
    true_prob_eff = true_prob_tox
  )
wt_sims
```

And under the BOIN12 design:

```{r}
b_model <- get_boin12(
  num_doses = 5, phi_t = 0.3, phi_e = 0.5, u2 = 40, u3 = 60, n_star = 6
) %>%
  dont_skip_doses() %>% 
  stop_at_n(n = 12) %>% 
  select_boin12_obd()
  
set.seed(2025)
b_sims <- b_model %>% 
  simulate_trials(
    num_sims = n_sim, 
    true_prob_tox = true_prob_tox,
    true_prob_eff = true_prob_tox
  )
b_sims
```

To compare designs, however, it is much more efficient to use the method presented in @sweeting2023 using `simulate_compare`:

```{r}
designs <- list(
  EffTox = et_model,
  WagesTait = wt_model,
  BOIN12 = b_model
)

sims <- simulate_compare(
  designs = designs, 
  num_sims = n_sim, 
  true_prob_tox = true_prob_tox,
  true_prob_eff = true_prob_tox
)
convergence_plot(sims)
```

We stress again, the design parameters in these examples have been chosen in haste merely for illustration. 
The work presented here in no way constitutes a valid comparison of these three designs.

For more information on running dose-finding simulations, refer to the simulation vignette.


# References
