---
title: "Treatment Combination Designs"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{Treatment Combination Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 5
)
```

# Introduction
The great majority of dose-finding designs are intended for use on a single treatment whose dose is being varied.
There are a small number of designs that are intended for dose-finding in multiple therapies combined, where the dose of more than one treatment can be varied.
A notable feature of this challenge is that the resulting dose space is multi-dimensional.
There is no longer a single optimal dose or maximum tolerable dose to identify but a contour of treatment combinations that have similar event probabilities.

# Outcome nomenclature
The outcome syntax used elsewhere in `escalation` to represent the patient-level doses administered and outcomes experienced has been extended to handle treatment combinations.
The only material change is in how the dose is written.
In a monotherapy setting, we would write `1` for dose-level 1 and `2` for dose-level 2, etc.
In the treatment combination setting, we use a dotted decimal notation (like that used in IP-addresses), so that, for example, the third dose-level of the first treatment paired with the fourth dose-level of the second treatment is written as `3.4`.
The same outcome designators `E`, `T`, `B` and `N` (representing efficacy only, toxicity only, both effiecacy and toxicity, and neither efficacy nor toxicity) are strung after the dose to represent the outcomes of patients in cohorts.
There currently are no designs implemented in `escalation` that investigate efficacy and toxicity in treatment combincations but the syntax can theoretically accomodate them.

For illustration, let us imagine a dose-finding trial of three treatments in a trial where each patient can either experience toxicity `T`, or no toxicity `N`.
Suppose further the first cohort of three patients is treated at the first dose-level of each of the three therapies and none of the patients experiences toxicity.
The outcome string to reflect this is:

`1.1.1NNN`

Now suppose, by way of some experimental design, we decide that the dose-level of the first treatment should be escalated to the second level whilst the dose-level of the second and third treatments is kept at level 1.
The dose string will become `2.1.1`.
Suppose we give this dose to a cohort of three new patients, and that two of the patients tolerate without toxicity but the third patient experiences tox.
The outcome string for the whole trial so far is:

`1.1.1NNN 2.1.1NNT`

This syntax can be used to describe the doses and outcomes for any arrangement of discrete dose-levels in any number of treatments.


# BOIN-COMB
The BOIN method was originally presented by @liu_bayesian_2015 for dose-finding in a single treatment.
BOIN-COMB [@lin2017bayesian] generalises BOIN to investigate doses in exactly two treatments (not more) in search of dose pairs with toxicity probability close to a pre-specified target rate.
Currently, BOIN-COMB is the only design for treatment combinations implemented in `escalation` 
We demonstrate some examples below.


# Implementation in `escalation`

## BOIN-COMB
We now demonstrate BOIN-COMB, focusing on how the shape of returned objects is different to monotherapy dose-finding.

In their treatment combination examples, the `BOIN` package authors investigate a situation where there are three doses of the first treatment being paired with five doses of the second treatment, targeting a 30% toxicity rate.
The first big difference compared to monotherapy dose-finding is that `num_doses` now must be a vector rather than a scalar integer:

```{r, message=FALSE}
library(escalation)

num_doses <- c(3, 5)
target <- 0.3
```

We create a dose selector:

```{r}
boin_fitter <- get_boin_comb(
  num_doses = num_doses, target = target
)
```

By default, the lower and upper toxicity rate thresholds in BOIN-COMB are set to `p.saf = 0.6 * target = 0.18` and `p.tox = 1.4 * target = 0.42`, so the above call is equivalent to:

```{r}
model <- get_boin_comb(
  num_doses = num_doses, target = target, 
  p.saf = 0.18, p.tox = 0.42
)
```

Naturally we can override those with our own values to make escalation harder or easier.

We now fit the model to outcomes.
The authors investigate outcomes:

```{r}
outcomes <- "1.1NNN 2.1NNNNNNT 2.2NNNNNT"
```

I.e. the first cohort of three is treated at the lowest dose combination with no tox experienced; the second cohort, comprising seven patients, is dosed at the second dose-level of the first treatment paired with the first dose-level of the second treatment, with one patient experiencing tox; and the third cohort, comprising six patients, is dosed at the second dose-level of the first treatment paired with the second dose-level of the second treatment, again with one patient experiencing tox.
As you can see, a lot of information is conveyed in 27 characters.

Let's fit the model:

```{r}
set.seed(2025)
fit <- model %>% fit(outcomes)
```

leading to advice to continue:

```{r}
fit %>% continue()
```

at dose `2.3`:

```{r}
fit %>% recommended_dose()
```

### Final dose selection
BOIN-COMB, like some other designs, selects the final dose differently to how it selects doses mid-trial.
To achieve this in `escalation`, we need an extra selector that will kick-in when the parent selector(s) have selected a non-NA dose but expressed `continue == FALSE`, i.e. signaled _the trial ends now but we are interested in a dose_.
If used, it will almost surely come last in the selector chain:

```{r}
model <- get_boin_comb(num_doses = num_doses, target = target)  %>%
   stop_at_n(n = 12) %>%
   select_boin_comb_mtd()

outcomes <- '1.1NNN 1.2NTN 1.2NNN 2.1TNT'
model %>% fit(outcomes) %>% recommended_dose()
```

In the above example, `stop_at_n` stopped the trial because the threshold sample size was met, and the underlying algorithm identified that at least one dose was worthy of selection.
At this juncture, `select_boin_comb_mtd` took over and applied the method described by the authors.
Whilst the underlying selector(s) were busy conducting the trial (`continue == TRUE`), `select_boin_comb_mtd` kept silent.


## Combination with other dose selectors
We demonstrated above how `get_boin_comb` works with the auxiliary dose-selector `stop_at_n`.
In a similar way, `get_boin_comb` will work with many other dose-selectors in `escalation`.

For instance, we might wish to stop the trial once we have reached a target sample size at recommended dose.
For instance, the following model wants to recommended dose `2.1`:

```{r}
set.seed(2025)
model <- get_boin_comb(num_doses = num_doses, target = target)  %>%
   stop_when_n_at_dose(n = 3, dose = "recommended")

outcomes <- '1.1NNN 2.1NTN'
x <- model %>% fit(outcomes) 
recommended_dose(x)
```

But it has reached the target sample size at that dose so it advocates stopping:

```{r}
continue(x)
```

Likewise, we could have specified a dose explicitly by the dose-indices:

```{r}
set.seed(2025)
model <- get_boin_comb(num_doses = num_doses, target = target)  %>%
   stop_when_n_at_dose(n = 3, dose = c(2, 1))

outcomes <- '1.1NNN 2.1NTN'
x <- model %>% fit(outcomes) 
recommended_dose(x)
continue(x)
```

There are many other dose-selectors in escalation.
PLease get in contact if you discover a combination that does not work and you are pretty sure that it should!


## Dose paths
Dose-paths for combinations are theoretically possible.
However, at present they are not implemented yet:

```{r, error=TRUE}
cohort_sizes <- c(3, 3)
paths <- model %>% 
  get_dose_paths(
    cohort_sizes = cohort_sizes, 
    previous_outcomes = '1NNN', next_dose = 2
  )
```

For more information on working with dose-paths, refer to the dose-paths vignette.


## Simulation
Simulations are implemented for combination designs but there are some necessary differences compared to monotherapy situations.
To illustrate, we will continue with the BOIN-COMB design already introduced, that will use a modest sample size:

```{r}
model <- get_boin_comb(num_doses = num_doses, target = target)  %>%
   stop_at_n(n = 12) %>%
   select_boin_comb_mtd()
```

Let us specify a plausible true dose-toxicity scenario:

```{r}
true_prob_tox <- matrix(c(0.1, 0.2, 0.3, 0.4, 0.5,
                          0.25, 0.35, 0.45, 0.55, 0.65,
                          0.3, 0.4, 0.5, 0.6, 0.7),
                        nrow = num_doses[1],
                        ncol = num_doses[2],
                        byrow = TRUE)
true_prob_tox
```

As with all designs in `escalation`, we perform simulations like this:

```{r}
set.seed(2025)
sims <- model %>%
  simulate_trials(num_sims = 10, true_prob_tox = true_prob_tox)
```

By default, we get some useful output:

```{r}
sims
```

However, we can dive deeper.
The same generics that can be invoked on the sims object in monotherapy scenarios can also be used in combinations studies.
However, some of the formats of the returned objects are different to reflect that dose is multi-dimensional.
We consider those generics each briefly now.

The number of patients in each simulated trial behaves as before: 

```{r}
num_patients(sims)
```


The number of doses is now a vector:

```{r}
num_doses(sims)
```

The set of possible dose-indices is a list, with each element being a vector on integers:

```{r}
dose_indices(sims)
```

That some information is more succinctly represented in dotted-decimal string notation:

```{r}
dose_strings(sims)
```

The recommended final dose in each iterate is available in string form:

```{r}
recommended_dose(sims, dose_string = TRUE)
```

or in matrix form, with simulated iterates in rows, treatment in columns, and matrix cells represending the dose-index of the relevant treatment:

```{r}
recommended_dose(sims, dose_string = FALSE)
```

The number of patients at each dose for a given trial is now a multi-dimensional array, and a 2*2 matrix in BOIN-COMB.
The same info for many trials for convenience is represented as a list, with one item for each simulated iterate:

```{r}
n_at_dose(sims)
```

When focusing just on the sample size at the recommended dose, the data is represented as a vector of integers:

```{r}
n_at_dose(sims, dose = "recommended")
```

That same information is returned by:

```{r}
n_at_recommended_dose(sims)
```

Likewise, the number of toxicity events seen at each dose combination in each ieterate is returned in a congruent form:

```{r}
tox_at_dose(sims)
```

The total number of toxicities seen over all doses in each iterate is simple to understand:

```{r}
num_tox(sims)
```

BOIN-COMB does not assess efficacy, but if it did, equivalent functions would return the counts (but merely return NA here:)

```{r}
eff_at_dose(sims) %>% head(2)
```

```{r}
num_eff(sims)
```

The probability of each dose being recommended is presented in a data-frame, with column `dose` containing a vector of dose indices for each dose (e.g. `c(1, 1)` for the pair of first dose), and `dose_string` comprising the same in dotted-decimal string format:

```{r}
prob_recommend(sims)
```

The probability of each dose being administered to a patient is available average over all simulated iterates:

```{r}
prob_administer(sims, method = 0)
```

or within in each iterate:

```{r}
prob_administer(sims, method = 1)
```

The trial duration is simple to understand:

```{r}
trial_duration(sims)
```

`summary` gives a data-frame summary with doses (and stopping without selecting a dose) as rows and much of the above information in columns:

```{r}
summary(sims)
```

`as_tibble` returns one row for each dose in each simulated iterate, offering a deep-dive into the simulated trials:

```{r}
tibble::as_tibble(sims)
```

Furthermore, the simulation comparison method of @sweeting2023 is implemented, allowing us to efficiently investigate small differences in competing designs with code like:

```{r}
model1 <- get_boin_comb(
  num_doses = num_doses, target = target, 
  p.saf = 0.6 * target, p.tox = 1.4 * target
) %>% 
  stop_at_n(n = 12) %>%
   select_boin_comb_mtd()

model2 <- get_boin_comb(
  num_doses = num_doses, target = target, 
  p.saf = 0.5 * target, p.tox = 1.5 * target
) %>% 
  stop_at_n(n = 12) %>%
   select_boin_comb_mtd()

designs <- list(
  model1 = model1,
  model2 = model2
)

set.seed(2025)
sims <- simulate_compare(
  designs = designs,
  num_sims = 10,
  true_prob_tox = true_prob_tox
)
convergence_plot(sims)
```

```{r}
tibble::as_tibble(sims)
```

For more information on running dose-finding simulations, refer to the simulation vignette.


# References
